---
title: 'Reproducible Research: Peer Assessment 1'
author: "Jim Callahan"
date: "September 13, 2015"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
This report analyzes the number of steps taken by an anonymous individual
user of a personal fitness armband device similar to the Nike "Fit" armband. 
The number of steps were measured over five minute intervals, 
24 hours a day during the months of October and November 2012.

The 2012 steps per five minute interval data is for the months of
October (30 days) and November (31 days) for a total of 61 days. 
In each hour there are 12 "five minute" intervals 
(60/5 = 12 intervals per hour). Thus, there are 288 five minute 
intervals in a 24 hour day (288 = 24 hours * 12 intervals per hour).
Therefore, with 288 measurements per day for 61 days (24 hour days) 
one would **expect 17,568 observations** (17,568 = 61 days * 288 per 24 hour day).


### Loading and preprocessing the data
The dataset is stored in a comma-separated-value (CSV) file in the main direcory of a **GitHub** repository. So, we can load the data with an R "read.csv()" function. In this case
name the **R** dataframe **"activity"**, the same name as the input filename. Finally, the initial struture of the **R** dataframe is shown with the R str() function:
```{r, echo=TRUE}
#### Set directory to the local GitHub project of this assignment.
setwd("~\\GitHub\\RepData_PeerAssessment1")

activity <- read.csv("activity.csv", 
                     na.strings = "NA", stringsAsFactors = FALSE )

str(activity)
```

**As expected**, for 61 days, **the "activity" data frame has 17,568 observations**.
The **"activity"** data frame has three variables: **"steps", "date"** and **"interval"**.
The initial values for the **"steps"** variable are missing and are marked as "NA".
The  **"date"** variable is a character string we will want to convert to 
an **R** date type using the **R** "as.Date()" function.
The **"interval"** variable is an integer that intially appears to be incremented
by 5 for each observation, this first impression will be modified on 
closer observation. 

So, let's convert **"date"** to an R date type and take a closer look 
at **"steps"** and **"interval"**. We want to know if all of the **"steps"** are missing?
or if not all are missing, how many are missing? and what percentage of the
dataset is that? For the **"interval"** variable, we want to know how often the pattern restarts
at zero (or does it increase all the way through the data set?).
We will print 289 observations of **"interval"**, one more than the 288 observations per day.
```{r, echo=TRUE}
activity$date <- as.Date(activity$date)   # coerce "date" to date data type
sum(is.na(activity$steps))                # How many missing values?
mean(is.na(activity$steps))               # What percent are missing values?
```

Over 2,300 observations of the **"steps"** variable are missing, 
while this is a lot; it is still only 13.1% of the 17,568 observations. 
So, for our initial analysis, we can simply ignore the missing values, 
by removing them and only work with complete cases. Later, we can try
to guess (technically, "impute") the missing values in a process
called "imputation" and see whether that changes the analysis. Not shown,
but there are no missing values for **"date"** and **"interval"**.

As for **"interval"**, What does a full daily cycle look like?
```{r, echo=TRUE}
head(activity$interval, 289)              # What does a full daily cycle look like?
```
The **"interval"** variable does not make sense as an integer.  
Although, the **"interval"** variable does reset to zero at observation 289,
as expected (recall there are 288 observations per 24 hour day); 
the 288th observation is 2,355 rather than the 1,435 one would expect
if one multiplied the intervals 0 (zero) through 287 by 5 (1,435 = 287*5). 
That is a big gap between 2,355 and 1,435 so something different is going on.  

If we examine the first dozen observations we see the **"interval"** variable jumps 
from 55 to 100. It is 100 when it should be 60. But, wait, if the "1" in "100" 
represents "one hour" and the "23" in "2355" represents "23 hours" then it is clear 
that the **"interval"** variable is actually hours and minutes with the leading zeros removed. 
That is, "100" should be understood as "01:00" and "2355" should be understood as "23:55" and so on.

We can fix the integer representation by using the **R "sprinf()"** function 
to restore the leading zeros to the time and store the result in a variable 
named **"HHMM"** which in turn, can be combined with the date to build a POSIX standard 
date time string in a variable named, **"datetime"**.
```{r, echo=TRUE}
# Convert the interval to HHMM by formating with leading zero
activity$HHMM <- sprintf("%04d",as.integer(activity$interval))
# Now we can combine date and time as a string
# and format the resulting string as a POSIX datetime string
datetimestring     <- paste(activity$date, activity$HHMM)
activity$datetime  <- strptime(datetimestring, 
                               "%Y-%m-%d %H%M", tz = "")

# Use the weekday() function to interograte the POSIX datatime string 
# to obtain a day of week abbrevation ("Sun", "Mon", "Tue", etc) for each date
activity$dayofweek <- weekdays(activity$datetime, abbreviate=TRUE)

# Create a weekday/weekend factor by first assigning all data to "weekday"
# and then check for "Sat" or "Sun" and reassign to "weekend".
activity$daytype = "weekday"
activity$daytype[activity$dayofweek == "Sat" | activity$dayofweek == "Sun"] <- "weekend"

# Convert to R "factor"" data type
activity$dayofweek = factor(activity$dayofweek)
activity$daytype   = factor(activity$daytype)
```
With these changes, the **"activity"** data set is ready for the first stage
of our analysis where we simply ignore (remove) the missing values. 
But, for cosmetic reasons, we might want to reorder the **"activity"** data frame variables 
in a more logical order:
```{r}
activity <- subset( activity, select = c(datetime, date, dayofweek, daytype, HHMM, interval, steps))
str(activity)
```


### What is mean total number of steps taken per day?
If we sum the steps for all of the intervals for a given date, we have a daily total.
This use of the **R aggregate()** function follows an example in Jared Lander's 
book *"R for Everyone"* page 121 where he uses the **"diamonds"** data frame 
which comes with the **ggplot2** package.

Once we have a daily total for each of the days; we can calcuate a value for
the average (mean) and median by removing the NAs, that would otherwise
cause an NA result.  We round the steps to zero because users expect a whole number
of steps and we already have five significant digits of precision.
```{r, echo=TRUE}
PerDay <- aggregate(steps ~ date, data=activity, sum)  
meanstepsperday <- round(mean(PerDay$steps, na.rm = TRUE), digits = 0)
meanstepsperday

medianstepsperday <- median(PerDay$steps, na.rm = TRUE) 
medianstepsperday
```

The **average (mean) number of steps per day** is 
**`r format(meanstepsperday, format="f", big.mark = ",")` **; 
and the **median number of steps per day** is very close at 
**`r format(medianstepsperday, format="f", big.mark = ",")`**.  

### What is the average daily activity pattern?
As a "sanity check", the 10,000+ average steps per day seems plausible 
for an active person with a 10,000 steps a day goal, which seems to be a popular goal:

> "The origins of the 10,000-steps recommendation aren't exactly scientific. Pedometers 
> sold in Japan in the 1960s were marketed under the name "manpo-kei," which translates 
> to "10,000 steps meter"  ...studies conducted since then suggest that people who 
> increased their walking to 10,000 steps daily experience health benefits."  
> Rachael Rettner, *"The Truth About '10,000 Steps' a Day"* 
LiveScience.org, March 2014  
retrieved from **LiveScience.org** during September 2015  
http://www.livescience.com/43956-walking-10000-steps-healthy.html

```{r, echo=FALSE}
hist(PerDay$steps, breaks=25,
     xlab = "Total Steps per Day",
     ylab = "Frequency (number of days)",
     main = paste("Histogram of Total Steps Per Day", 
                  "\n Mean = " , round(mean(PerDay$steps), digits=0),
                  "\n Median =", median(PerDay$steps) 
                  
                  ) 
     )
# Only last line prints because values are so close together.
abline(v = mean(PerDay$steps), col = "Green")
abline(v = median(PerDay$steps), col = "Blue")
```
While a central value (mean or median) near 10,000 seems plausible for an active 
person with a 10,000 steps per day goal; the extremes of near zero steps per day 
and a maximum over 20,000 steps per day may require further inquiry. For example, 
did the person spend a sick day in bed (with near zero steps)? and did the person 
participate in a 10,000 step walk in addition to their normal 10,000 steps  
(resulting in over 20,000 steps per day)?  

What interval contains the maximum number of steps? and how does the number of steps
vary over the course of an average day?
```{r}
# Calculate steps per five minute interval
# (in 24 hour cycle)
PerIntervalMean    <- aggregate(steps ~ factor(HHMM), activity, mean)
PerIntervalMedian  <- aggregate(steps ~ factor(HHMM), activity, median)
ColumnNames <- c("HHMM", "steps")
# colnames(PerIntervalSum)    <- ColumnNames
colnames(PerIntervalMean)   <- ColumnNames
colnames(PerIntervalMedian) <- ColumnNames


PerIntervalMean$timeofday    <- strptime(PerIntervalMean$HHMM, "%H%M", tz = "")
PerIntervalMedian$timeofday  <- strptime(PerIntervalMedian$HHMM, "%H%M", tz = "")

Max5MinuteSteps <- PerIntervalMean[PerIntervalMean$steps == max(PerIntervalMean$steps), ]
Max5MinuteSteps
```
The interval beginning at 8:35 AM on average has the most steps with 
`r round(max(PerIntervalMean$steps), digits=0)` steps.

```{r}
plot(PerIntervalMean$timeofday, PerIntervalMean$steps, type = "l",
    xlab = "Starting time of each five minute interval",
    ylab = "Avg Steps per Interval",
    main = paste("Average Steps Per Five Minute Interval", 
                 "\n Maximum = " , round(max(PerIntervalMean$steps), digits=0),
                 "\n Maximum occurs at: ", format(Max5MinuteSteps$timeofday, "%H:%M AM"),
                 "Interval"
                 )
    )

abline(h = round(max(PerIntervalMean$steps), digits=0), col = "red") 
```
The number of steps in the graph could be interpreted as the person sleeps 
between midnight and 5 AM, gets up around 5 or 6 AM and commutes to work by 9 AM, 
has a lunch hour and returns home between 5 PM and 7 PM. Such a pattern would
be consistent with the peaks, but is conjecture and the reality may be different.


### Imputing missing values
So far, we have ignored (omitted) the missing values, and as noted earlier, 
over 2,300 observations of the **"steps"** variable are missing, while this is a lot; 
it is still only 13.1% of the 17,568 observations. 
```{r, echo=TRUE}
sum(is.na(activity$steps))       # How many missing values?  
mean(is.na(activity$steps))      # What percent missing values?
```

There seems to be a regular pattern for time of day, but we still have to decided
whether we should impute with 5 minute interval **averages** or **medians**.

We have already made a graph of the **five minute means (averages)**, and for comparison 
here is a graph of **median steps per five minutes**, it peaks at 60 steps
and often takes on a zero value.

```{r}
plot(PerIntervalMedian$timeofday, PerIntervalMedian$steps, type = "l",
          main = paste("Median Steps Per Five Minute Interval", 
                       "\n Maximum = " , round(max(PerIntervalMedian$steps), digits=0)
                       )

    )
```

Overall, the median graph shows a much sharper, commute to work, lunch hour and return
from work and sleep pattern. It is much less noisey and extreme. So, let's use the **"steps"** variable in the **"PerIntervalMedian"** data frame to create a variable we can use to **impute (fill-in) the missing values** of the **"steps"** variable in the **"activity"** data frame. We need to expand the **"PerIntervalMedian"** version of **"steps"** from just one day 
to all **61 days** (17,568 observations). We can do this in **R** by repeating 
the variable **61 times** using the **R rep()** function:

```{r}
# Expand PerIntervalMedian from one day to 61 days by repeating the daily values
activity$fill <- rep(PerIntervalMedian$steps, 61)
str(activity$fill)
```
We now have a prediction variable **"fill"** that we can use to "fill-in" (replace)
the missing values (NA) of the **"steps"** variable.

The strategy for replacement is to "zero-out" the unwanted values in **"steps"** 
and **"fill"** using **R's** subsetting (square brackets) and the is.na() function.
This way we can zero out the NA's in a copy of **"steps"** (we want to get rid of) 
and zero out the opposite values in a copy of **"fill" ** that correspond to non-NA 
values in **"steps"** (we want to keep). Once we have the zeros in the correct places 
we can simply add the variables together into a third variable.

```{r}
# Keep the non-NA values; zero-out the NA observations.
stepsNA2Zero <- activity$steps
stepsNA2Zero[is.na(activity$steps)] <- 0

# Keep the values needed to fill NAs; but zero-out the others.
# NOTE: this subset is for "not-NA"
filltheNAs <- activity$fill
filltheNAs[!is.na(activity$steps)] <- 0

# Now we can add the values
activity$stepsNoNA <- stepsNA2Zero + filltheNAs
activity$stepsNoNA <- round(activity$stepsNoNA, digits = 0)
```

We confirm our NA fill worked, and the number of NAs in **"activity$stepsNoNA"** is zero.

```{r}
sum(is.na(activity$stepsNoNA))       # How many missing values in "NoNA" version?
```


### Are there differences in activity patterns between weekdays and weekends?
Using our newly filled in **"stepsNoNA"** variable we can look to see if there is any 
difference between weekdays and weekends.
```{r}
ColumnNames <- c("HHMM", "daytype", "stepsNoNA")
DayTypePerIntervalMean    <- aggregate(stepsNoNA ~ factor(HHMM)+daytype, data=activity, mean)
colnames(DayTypePerIntervalMean) <- ColumnNames
DayTypePerIntervalMean$timeofday    <- strptime(DayTypePerIntervalMean$HHMM, "%H%M", tz = "")
DayTypePerIntervalMean$interval <- as.integer(DayTypePerIntervalMean$HHMM)

WeekdayPerIntervalMean   <- DayTypePerIntervalMean[DayTypePerIntervalMean$daytype == "weekday", ]
WeekendPerIntervalMean   <- DayTypePerIntervalMean[DayTypePerIntervalMean$daytype == "weekend", ]

require(ggplot2)
# use facets to display type.
# HONOR CODE: "R Graphics Cookbook" by Winston Chang page 163 and 208
par(mar = c(4, 4, 4, 1) )
line2 <- qplot( data = DayTypePerIntervalMean, x = interval, y = stepsNoNA) +
    geom_line() + facet_grid(. ~daytype) +  
    ggtitle("Mean Steps per Five Minute Interval") +  
    xlab("Five minute interval number - 0 to 287") +  
    ylab("Avg Steps per Interval")
line2
```
The weekend has activity throughout the day and is less peaked at typical commuter hours.
